---
title: "DAPC all data"
author: "Wandrille Duchemin"
date: "2025-03-05"
output: html_document
---


```{r}
library(adegenet)
library(pegas)
```

## reading the data

```{r}
vcfFile = '053_removing_siblings/FM_0.70_mD_2_MD_30_FMi_0.35_LD_thin.siblings_removed.vcf.gz'

# reading files in adegenet format: https://cran.r-project.org/web/packages/pegas/vignettes/ReadingFiles.pdf
loci_data = read.vcf( vcfFile , to=20000 )
genind_data = loci2genind(loci_data)
genind_data
```

adding population data 
```{r}
pop_table = read.table('/scicore/home/muellepi/GROUP/albopictus/info/albo_global_pops_UNSD.txt')
colnames( pop_table ) = c('indID','population','nativeInvaded')
rownames( pop_table ) =  paste( pop_table$indID , pop_table$indID , sep='_' )


pop(genind_data) = pop_table[ rownames( genind_data@tab  ) , 'population' ]

table( pop(genind_data) )
```

```{r}
## also preparing the table with the samples in the expected order for further selection of native/invasive
pop_table = pop_table[ rownames( genind_data@tab  ) ,  ]
```

## DAPC with all data

adegenet DAPC tutorial https://raw.githubusercontent.com/thibautjombart/adegenet/master/tutorials/tutorial-dapc.pdf

```{r}
#grp <- find.clusters(genind_data, max.n.clust=50)
# keep 417 PCs (all)
# keep 3 clusters (min BIC)
```


```{r}
grp <- find.clusters(genind_data, max.n.clust=50, n.pca = 417, n.clust = 3)
grp$size
```

```{r}
table(pop(genind_data), grp$grp)
```

Indonesia is in its own cluster,

```{r}
unique( pop(genind_data)[ grp$grp == 1 ] )
```

cluster2:
```{r}
unique( pop(genind_data)[ grp$grp == 2 ] )
```
cluster3:
```{r}
unique( pop(genind_data)[ grp$grp == 3 ] )
```
in both cluster 2 and 3:
```{r}
a = unique( pop(genind_data)[ grp$grp == 2 ] ) 
b = unique( pop(genind_data)[ grp$grp == 3 ] ) 
intersect(a,b)
```

in both cluster 1 and 2:
```{r}
a = unique( pop(genind_data)[ grp$grp == 1 ] ) 
b = unique( pop(genind_data)[ grp$grp == 2 ] ) 
intersect(a,b)
```

in both cluster 1 and 3:
```{r}
a = unique( pop(genind_data)[ grp$grp == 1 ] ) 
b = unique( pop(genind_data)[ grp$grp == 3 ] ) 
intersect(a,b)
```


```{r}

dapc1 <- dapc( genind_data , grp$grp , n.pca = 100 , n.da = 2)
scatter(dapc1)
```

```{r}
scatter(dapc1, scree.da=FALSE, bg="white", pch=20, cell=0, cstar=0, 
        cex=1,clab=0, leg=TRUE, txt.leg=paste("Cluster",1:3))
```

Let's save that information somewhere

```{r}

## I simplify the assigned posterior probability 
df = cbind( dapc1$ind.coord , round( dapc1$posterior , 3 ) )
write.csv(df,"071_DAPC_all.3clusters.csv", row.names = TRUE)
```

```{r}
## alternative with less PCs
#grp <- find.clusters(genind_data, max.n.clust=50, n.pca = 100, n.clust = 7 )
#dapc1 <- dapc( genind_data , grp$grp , n.pca = 20, n.da = 6 )

#scatter(dapc1, posi.da="bottomright", bg="white",
#        pch=17:22, cstar=0,  scree.pca=TRUE,
#        posi.pca="bottomleft")
```


## DAPC without the indonesian samples


```{r}

sub = genind_data[ pop(genind_data) != "Indonesia" , ]

grp_sub = find.clusters( sub , n.pca = 200 , n.clust = 4)
dapc_sub <- dapc( sub , grp_sub$grp , n.pca = 100 , n.da = 4)
scatter(dapc_sub)
```
Now the "excluded" one is Vanuatu

```{r}
table(pop(sub),grp_sub$grp)
```

```{r}
## I simplify the assigned posterior probability 
df = cbind( dapc_sub$ind.coord , round( dapc_sub$posterior , 3 ) )
write.csv(df,"071_DAPC_noIndonesia.4clusters.csv", row.names = TRUE)
```

## no Indonesia or Vanuatu

```{r}

sub = genind_data[ !(pop(genind_data) %in% c("Indonesia" , "Vanuatu" ) )  , ]

grp_sub = find.clusters( sub , n.pca = 200 , n.clust = 3)
dapc_sub <- dapc( sub , grp_sub$grp , n.pca = 100 , n.da = 2)
scatter(dapc_sub)
```


```{r}
## getting assignments with Vanuatu (but not Indonesia)
df =read.csv("071_DAPC_noIndonesia.4clusters.csv" , row.names = 1)
cl = apply( df[ rownames( sub@tab  ) ,c('X1' , 'X2' , 'X3' , 'X4' ) ]  , 1 ,which.max )
# comparing with the current:
table(  grp_sub$grp , cl )
```
So this does not really bring much more.

## invaded 

```{r}
sub_invaded = genind_data[ pop_table$nativeInvaded == "invaded" , ]

grp_sub_invaded = find.clusters( sub_invaded , n.pca = 200 , n.clust = 2)
dapc_sub_invaded <- dapc( sub_invaded , grp_sub_invaded$grp , n.pca = 100 , n.da = 1)
scatter(dapc_sub_invaded)
```

```{r}
## getting assignments with all to compare
df =read.csv("071_DAPC_all.3clusters.csv" , row.names = 1)
cl = apply( df[ rownames( sub_invaded@tab  ) ,c('X1' , 'X2' , 'X3'  ) ]  , 1 ,which.max )
# comparing with the current:
table(  grp_sub_invaded$grp , cl )
```
We find the same groupings

```{r}
# compare with teh 4 custers when we remve indonesia
df =read.csv("071_DAPC_noIndonesia.4clusters.csv" , row.names = 1)

cl = apply( df[ rownames( sub_invaded@tab  ) ,c('X1' , 'X2' , 'X3' , 'X4'  ) ]  , 1 ,which.max )
# comparing with the current:
table(  grp_sub_invaded$grp , cl )
```
Cluster 2 is not the one I expected to see keeping up here...

## native 


```{r}
sub_native = genind_data[ pop_table$nativeInvaded == "native" , ]

grp_sub_native = find.clusters( sub_native , n.pca = 100 , n.clust = 2)
table( grp_sub_native$grp , pop(sub_native) )
```
as expected, Indonesia vs. all the rest is the dominating signal


## native without indonesia

```{r}
sub_native = genind_data[ (pop_table$nativeInvaded == "native") & (pop_table$population != "Indonesia") , ]

grp_sub_native = find.clusters( sub_native , n.pca = 60 , n.clust = 1)
table( grp_sub_native$grp , pop(sub_native) )
```

This supports having a single cluster


## recap

clustering of all data: 3 clusters, one being Indonesia
When we remove indonesia: 4 clusters 
Invaded or native separated: nothing new

```{r}
df_A =read.csv("071_DAPC_all.3clusters.csv" , row.names = 1)
df_noI =read.csv("071_DAPC_noIndonesia.4clusters.csv" , row.names = 1)
cl_noI = apply( df_noI[  ,c('X1' , 'X2' , 'X3' , 'X4' ) ]  , 1 ,which.max )
cl_A = apply( df_A[ row.names(df_noI)  ,c('X1' , 'X2' , 'X3' ) ]  , 1 ,which.max )
# comparing with the current:
table( cl_noI , cl_A )
```
```{r}
NI = pop_table[ row.names(df_noI) , 'nativeInvaded' ] 
PO = pop_table[ row.names(df_noI) , 'population'    ] 

table( NI , cl_noI  )
```
```{r}
for(c in 1:4){
  
  print( paste("cluster",c))
  print( table( NI[ cl_noI == c ]  , PO[ cl_noI == c ]  )   )

}

```

```{r}
cluster_countries = list()
for( i in 1:4){
  cluster_countries[[i]]= unique( PO[ cl_noI == i ] )  
}

for( a in 1:4){
  for( b in 1:4){
    if(a<b){
      I = intersect( cluster_countries[[a]] , cluster_countries[[b]] )     
      if(length(I)>0){
        print(paste('common between',a,'and',b,':',paste(I,collapse=' ; ')))
      }
    }
  }
}

```

## DAPC to test existing clustering 


### native invaded 

```{r}
dapc1 <- dapc( genind_data , pop_table$nativeInvaded , n.pca = 100 , n.da = 1)
scatter(dapc1)
```
It is apparently able to find a good discriminant function

```{r}
#round( dapc1$posterior , 3 )
#assignplot(dapc1)
cl =  c('invaded_post','native_post')[ apply( dapc1$posterior  , 1 ,which.max ) ]
table( cl , pop_table$nativeInvaded )
```

3 natives seen as invaded

```{r}
pop_table$population[ (cl=='invaded_post' & pop_table$nativeInvaded == 'native') ]
```
### population

```{r}
dapc1 <- dapc( genind_data , pop(genind_data) , n.pca = 200 , n.da = 7)
scatter(dapc1)
```

Let's remove Indonesia

```{r}
sub = genind_data[ pop(genind_data) != "Indonesia" , ]

dapc1 <- dapc( sub , pop(sub) , n.pca = 150 , n.da = 6)
scatter(dapc1)
```

Let's remove Vanuatu as well
```{r}
sub = genind_data[ !(pop(genind_data) %in% c("Indonesia",'Vanuatu')) , ]

dapc1 <- dapc( sub , pop(sub) , n.pca = 150 , n.da = 8)
scatter(dapc1)
#scatter(dapc1, scree.da=FALSE, bg="white", pch=20, cell=0, cstar=0, 
#        cex=1,clab=0, leg=TRUE)
```

Remove Phillipines and Fiji and ChristmasIsland ?
```{r}
sub = genind_data[ !(pop(genind_data) %in% c("Indonesia",'Vanuatu','Philippines','Fiji','ChristmasIsland')) , ]

dapc1 <- dapc( sub , pop(sub)  , n.pca = 150 , n.da = 5)
scatter(dapc1)
#scatter(dapc1, scree.da=FALSE, bg="white", pch=20, cell=0, cstar=0, 
#        cex=1,clab=0, leg=TRUE)
```

Japan seems to group differently (closer to south-east asian places) than in the k-mean clustering

---

From DAPC tutorial:

'''
Besides scatterplots of discriminant functions, group memberships of DAPC can be exploited.
Note that caution should be taken when interpreting group memberships of a DAPC based
on too many PCs, as there are risks of overfitting the discriminant functions (see section
below). But despite this possible bias, group memberships can be used as indicators of how
clear-cut genetic clusters are. Note that this is most useful for groups defined by an external
criteria, i.e. defined biologically, as opposed to identified by k-means. It is less useful for
groups identified using find.clusters, since we expect k-means to provide optimal groups
for DAPC, and therefore both classifications to be mostly consistent.
'''

We could do this for populations eventually, although I am not sure exactly what we would learn from it...

